Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks

### 2025-02-04 report/emse24-revised.tex
feat: init emse24-revised.tex

Figure~\ref{fig:data-collection} illustrates the data collection process employed in this study to gather Jupyter notebooks written in Python from GitHub and Kaggle. We used GitHub's advanced search syntax\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}, with the following search query: \lstinline[language={}]$language:"Jupyter Notebook"$ to mine 49.9 thousand Jupyter notebooks from public repositories on GitHub\footnote{Collected on June 22, 2023}. We used the pre-existing dataset KGTorrent~\citep{quaranta2021kgtorrent} for Kaggle since it does not support advanced code-based search like GitHub. To the best of our knowledge, KGTorrent is the largest dataset of Python Jupyter notebooks obtained from Kaggle, consisting of 247.9 thousand notebooks. Consequently, we started with 283 GB of data comprising 297.8 thousand Python Jupyter notebooks.

@@

@@

>>>>

Figure~\ref{fig:data-collection} illustrates the data collection process employed in this study to gather Jupyter notebooks written in Python from GitHub and Kaggle. GitHub is a popular platform to host and collaborate on software projects. It offers extensive tools for project management such as issue tracking, pull requests, code reviews and tooling for continuous integration and delivery. In contrast, Kaggle is focused specifically on data science and machine learning. It provides integrated computational resources, a rich repository of datasets, and a platform to host machine learning competitions. While GitHub supports formal project workflows, Kaggle encourages a more informal approach to data exploration and modeling, making it a preferred choice for data science practitioners looking for hands-on experience with developing machine learning models.

We used GitHub's advanced search syntax\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}, with the following search query: \lstinline[language={}]$language:"Jupyter Notebook"$ to mine 49.9 thousand Jupyter notebooks from public repositories on GitHub\footnote{Collected on June 22, 2023}. We used the pre-existing dataset KGTorrent~\citep{quaranta2021kgtorrent} for Kaggle since it does not support advanced code-based search like GitHub. To the best of our knowledge, KGTorrent is the largest dataset of Python Jupyter notebooks obtained from Kaggle, consisting of 247.9 thousand notebooks. Consequently, we started with 283 GB of data comprising 297.8 thousand Python Jupyter notebooks.

@@

\subsection{Descriptive Statistics}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{notebook-size-distribution.pdf}
  \caption{Number of notebooks in each category of...}
  \label{fig:notebook-size-distribution}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{cell-per-notebook-size-distribution.pdf}
  \caption{Distribution of number of cells for various notebook sizes.}
  \label{fig:cell-per-notebook-size-distribution.pdf}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{fms-per-notebook-size-distribution.pdf}
  \caption{Distribution of assertions, prints and last cell statements for various notebook sizes.}
  \label{fig:fms-per-notebook-size-distribution}
\end{figure}

% TODO should we add the descriptive stats as a table?
% TODO do we need to extrapolate more from figure 5?
Figure~\ref{fig:notebook-size-distribution} shows the distribution of the number of code cells for all notebooks analyzed in this study. The distribution is grouped into three categories---namely small, medium and large---based on the first and third quartiles. This is done since the data is large, non-uniform and contains many outliers. Figure~\ref{fig:cell-per-notebook-size-distribution.pdf} further magnifies the distribution of the number of cells within each category. The dataset mostly contains medium size notebooks (9 to 26 code cells), followed by small (1 to 8 code cells) and large (27 to 772 code cells) size notebooks.

Figure~\ref{fig:fms-per-notebook-size-distribution} shows the distribution of assertions, prints and last cell statements across the different notebook size categories. A logarithmic scale is used for the count since the data is large and contains outliers. Overall, the visualization suggests an increase in the use of all feedback mechanisms as notebook size grows. Print statements are the most frequently used across all notebook sizes.

@@

\subsection{Complexity of Notebooks}

Analyzing the feedback mechanisms with respect to the complexity of Jupyter notebooks remains a promising direction for future research. We believe that the primary challenge here will be to develop an appropriate definition for notebook complexity.

To the best of our knowledge, no formal definition or metrics to quantify notebook complexity exist to date. In their research, \citet{grotov2022large-scale} investigated the complexity of Python code written inside Jupyter notebooks using a subset of 15 metrics derived from prior literature. These metrics however were primarily developed for code structured in an object-oriented programming paradigm. Consequently, the applicability of these metrics to code within notebooks is questionable, particularly given that research indicates notebooks frequently diverge from established software engineering best practices \cite{pimentel2019large-scale}.

It is further unclear how metrics such as the size of the training data, complexity of the learning algorithm, computational costs and CO\(_2\) emissions associated with training can be incorporated into the definition for notebook complexity. Addressing these challenges would facilitate the stratification of our dataset into distinct categories---namely low, medium, and high complexity---which can enhance the depth of this study by enabling a more nuanced analysis of feedback mechanisms within this context.
